---
title: "soundscapes-acoustic-data-prep"
format: html
editor: visual
---

## Filtering the data

### Urchin Barren Sites

### - Harris Island, HydroMoth 12

```{r}
#reading in the data
library(here)
harris_island <- read.csv("data/barren_sites/harris_island_hydromoth12.csv")
```

### Filtering to sync time duration across forest and barren sites

```{r}
library(dplyr)
library(lubridate)

# Define your range
start_datetime <- ymd_hms("2025-07-23 23:59:43", tz = "UTC")
end_datetime   <- ymd_hms("2025-07-28 23:59:32", tz = "UTC")

filtered_harris_island <- harris_island %>%
  # 1. Remove any leading/trailing whitespace from the columns
  mutate(
    DATE..UTC. = trimws(DATE..UTC.),
    TIME..UTC. = trimws(TIME..UTC.)
  ) %>%
  # 2. Combine and convert, silencing the 'parse failed' warning
  mutate(
    full_datetime = ymd_hms(
      paste(DATE..UTC., TIME..UTC.), 
      tz = "UTC", 
      truncated = 3, 
      quiet = TRUE
    )
  ) %>%
  # 3. Filter out the NAs (the 112 failed rows) and apply your date range
  filter(!is.na(full_datetime)) %>%
  filter(between(full_datetime, start_datetime, end_datetime))

# Check the final count
nrow(filtered_harris_island)


```

### - Samsing Cove, Hydromoth 6

#reading in the data

library(here)

samsing_cove \<- read.csv("data/barren_sites/samsing_cove_hydromoth6.csv")

```{r}
#Reading in the data
library(here)
samsing_cove <- read.csv("data/barren_sites/samsing_cove_hydromoth6.csv")
```

### Filtering to sync time duration across forest and barren sites

```{r}
library(dplyr)
library(lubridate)

# Define your range
start_datetime <- ymd_hms("2025-07-23 23:59:43", tz = "UTC")
end_datetime   <- ymd_hms("2025-07-28 23:59:32", tz = "UTC")

filtered_samsing_cove <- samsing_cove %>%
  # 1. Remove any leading/trailing whitespace from the columns
  mutate(
    DATE..UTC. = trimws(DATE..UTC.),
    TIME..UTC. = trimws(TIME..UTC.)
  ) %>%
  # 2. Combine and convert, silencing the 'parse failed' warning
  mutate(
    full_datetime = ymd_hms(
      paste(DATE..UTC., TIME..UTC.), 
      tz = "UTC", 
      truncated = 3, 
      quiet = TRUE
    )
  ) %>%
  # 3. Filter out the NAs (the 112 failed rows) and apply your date range
  filter(!is.na(full_datetime)) %>%
  filter(between(full_datetime, start_datetime, end_datetime))

# Check the final count
nrow(filtered_samsing_cove)

```

### - Sandy Cove, HydroMoth 15

```{r}
#reading in the data
library(here)
sandy_cove <- read.csv("data/barren_sites/sandy_cove_hydromoth15.csv")
```

### Filtering to sync time duration across forest and barren sites

```{r}
library(dplyr)
library(lubridate)

# Define your range
start_datetime <- ymd_hms("2025-07-23 23:59:43", tz = "UTC")
end_datetime   <- ymd_hms("2025-07-28 23:59:32", tz = "UTC")

filtered_sandy_cove <- sandy_cove %>%
  # 1. Remove any leading/trailing whitespace from the columns
  mutate(
    DATE..UTC. = trimws(DATE..UTC.),
    TIME..UTC. = trimws(TIME..UTC.)
  ) %>%
  # 2. Combine and convert, silencing the 'parse failed' warning
  mutate(
    full_datetime = ymd_hms(
      paste(DATE..UTC., TIME..UTC.), 
      tz = "UTC", 
      truncated = 3, 
      quiet = TRUE
    )
  ) %>%
  # 3. Filter out the NAs (the 112 failed rows) and apply your date range
  filter(!is.na(full_datetime)) %>%
  filter(between(full_datetime, start_datetime, end_datetime))

# Check the final count
nrow(filtered_sandy_cove)
```

### Macro Kelp Forest Sites

### - Battery Island, HydroMoth 0

```{r}
#Reading in the data
library(here)
battery_island <- read.csv("data/macro_forest_sites/battery_island_hydromoth0.csv")
```

### Filtering to sync time duration across forest and barren sites

```{r}
library(dplyr)
library(lubridate)

# Define your range
start_datetime <- ymd_hms("2025-07-23 23:59:43", tz = "UTC")
end_datetime   <- ymd_hms("2025-07-28 23:59:32", tz = "UTC")

filtered_battery_island <- battery_island %>%
  # 1. Remove any leading/trailing whitespace from the columns
  mutate(
    DATE..UTC. = trimws(DATE..UTC.),
    TIME..UTC. = trimws(TIME..UTC.)
  ) %>%
  # 2. Combine and convert, silencing the 'parse failed' warning
  mutate(
    full_datetime = ymd_hms(
      paste(DATE..UTC., TIME..UTC.), 
      tz = "UTC", 
      truncated = 3, 
      quiet = TRUE
    )
  ) %>%
  # 3. Filter out the NAs (the 112 failed rows) and apply your date range
  filter(!is.na(full_datetime)) %>%
  filter(between(full_datetime, start_datetime, end_datetime))

# Check the final count
nrow(filtered_battery_island)
```

### - Crow Island, HydroMoth 0.5

```{r}
#Reading in the data
library(here)
crow_island <- read.csv("data/macro_forest_sites/crow_island_hydromoth0.5.csv")
```

### Filtering to sync time duration across forest and barren sites

```{r}
library(dplyr)
library(lubridate)

# Define your range
start_datetime <- ymd_hms("2025-07-23 23:59:43", tz = "UTC")
end_datetime   <- ymd_hms("2025-07-28 23:59:32", tz = "UTC")

filtered_crow_island <- crow_island %>%
  # 1. Remove any leading/trailing whitespace from the columns
  mutate(
    DATE..UTC. = trimws(DATE..UTC.),
    TIME..UTC. = trimws(TIME..UTC.)
  ) %>%
  # 2. Combine and convert, silencing the 'parse failed' warning
  mutate(
    full_datetime = ymd_hms(
      paste(DATE..UTC., TIME..UTC.), 
      tz = "UTC", 
      truncated = 3, 
      quiet = TRUE
    )
  ) %>%
  # 3. Filter out the NAs (the 112 failed rows) and apply your date range
  filter(!is.na(full_datetime)) %>%
  filter(between(full_datetime, start_datetime, end_datetime))

# Check the final count
nrow(filtered_crow_island)
```

### - Middle Island, HydroMoth 2

```{r}
#Reading in the data
library(here)
middle_island <- read.csv("data/macro_forest_sites/middle_island_hydromoth2.csv")
```

### Filtering to sync time duration across forest and barren sites

-   choosing the same start and end time across forest and barren sites to sync the data

```{r}
library(dplyr)
library(lubridate)

# Define your range
start_datetime <- ymd_hms("2025-07-23 23:59:43", tz = "UTC")
end_datetime   <- ymd_hms("2025-07-28 23:59:32", tz = "UTC")

filtered_middle_island <- middle_island %>%
  # 1. Remove any leading/trailing whitespace from the columns
  mutate(
    DATE..UTC. = trimws(DATE..UTC.),
    TIME..UTC. = trimws(TIME..UTC.)
  ) %>%
  # 2. Combine and convert, silencing the 'parse failed' warning
  mutate(
    full_datetime = ymd_hms(
      paste(DATE..UTC., TIME..UTC.), 
      tz = "UTC", 
      truncated = 3, 
      quiet = TRUE
    )
  ) %>%
  # 3. Filter out the NAs (the 112 failed rows) and apply your date range
  filter(!is.na(full_datetime)) %>%
  filter(between(full_datetime, start_datetime, end_datetime))

# Check the final count
nrow(filtered_middle_island)
```

## Combining into one data frame for the SVM

-   to correctly run the SVM, all data need to be in one dataframe

```{r}
all_data <- rbind(
  transform(samsing_cove,  site = "samsing_cove",  TYPE = "Barren"),
  transform(sandy_cove,    site = "sandy_cove",    TYPE = "Barren"),
  transform(harris_island, site = "harris_island", TYPE = "Barren"),
  transform(crow_island,   site = "crow_island",   TYPE = "Macro_Forest"),
  transform(middle_island, site = "middle_island", TYPE = "Macro_Forest"),
  transform(battery_island,site = "battery_island",TYPE = "Macro_Forest")
)

all_data$TYPE <- factor(all_data$TYPE, levels = c("Barren", "Macro_Forest"))
all_data$site <- factor(all_data$site)

```

## 
