---
title: "soundscapes-ecological-data-prep"
format: html
editor: visual
---

## Tidying the data

### UPC data

### - transforming forest and barren sites to be in same format

Transforming forested sites to long format

```{r}
#reading in forest upc data
library(here)
upc_forests <- read.csv("/Users/aminashenasa/soundscapes_data/data/ecological_data/Summer_Soundscapes_UPC_Forests - data_filtered.csv")

#checking column names
names(upc_forests)

```

Forested sites were surveyed in 5 meter segments while barren sites were surveyed in 10m segments, here I will be standardizing the two and converting the segments from 5 meters to 10 in forest sites.

```{r}
library(tidyverse)

# convert wide 5m segments to long 10m segments
old_long_upc_forests <- upc_forests %>%
  
  # rename 5 m segment column
  rename(SEGMENT5 = `SEGMENT.OF.THE.TRANSECT`) %>%
  
  # pivot all point columns to long format
  pivot_longer(
    cols = matches("SUBSTRATE|RELIEF|COVER|SUPERLAYER"), # include SUPERLAYER if present
    names_to = "name",
    values_to = "CLASSCODE"
  ) %>%
  
  # define CATEGORY from column name
  mutate(
    CATEGORY = str_extract(name, "SUBSTRATE|RELIEF|COVER|SUPERLAYER")
  ) %>%
  
  # remove empty points
  filter(!is.na(CLASSCODE), CLASSCODE != "") %>%
  
  # keep all original columns
  ungroup() %>%
  
  # assign 10 m segments by grouping adjacent 5 m segments
  mutate(
    SEG_START = as.numeric(str_extract(SEGMENT5, "^[0-9]+")),
    SEGMENT = case_when(
      SEG_START < 10 ~ "0-10m",
      SEG_START < 20 ~ "10-20m",
      TRUE ~ "20-30m"
    )
  ) %>%
  
  # calculate COUNT per 10 m segment while keeping metadata
  group_by(
    YEAR, MONTH, DAY,
    TIME, SEASON, SITE, TYPE, OBSERVER, BUDDY, TRANSECT,
    HEADING.OUT, `Start.Depth..0.meter.of.transect.tape..m..`,
    `End.Depth..30.meter.of.transect.tape...m..`,
    SEGMENT,
    CATEGORY,
    CLASSCODE
  ) %>%
  summarise(COUNT = n(), .groups = "drop") %>%
  
  # sort for easier inspection
  arrange(
    YEAR, MONTH, DAY,
    SITE, TRANSECT,
    SEGMENT,
    CATEGORY
  )
# saving file to cleaned_data folder
write_csv(old_long_upc_forests, "/Users/aminashenasa/soundscapes_data/data/ecological_data/cleaned_data/long_upc_forests.csv")

#tidying CLASSCODE column and adding in NAME column

```

Renaming barren file to match long forest file

```{r}
library(here)
#renaming file
long_upc_barrens <- read.csv("/Users/aminashenasa/soundscapes_data/data/ecological_data/Summer_Soundscapes_UPC_barrens - data_filtered.csv")

# saving file to cleaned_data folder
write_csv(long_upc_barrens, "/Users/aminashenasa/soundscapes_data/data/ecological_data/cleaned_data/long_upc_barrens.csv")
```

### Algae and Invert Swath

```{r}
full_algae_invert_swath <- read.csv("/Users/aminashenasa/soundscapes_data/data/ecological_data/Summer_Soundscapes_Algae_Invert_data.csv")

# Filter to only BARREN and MACRO, excluding Samsing Forest
filtered_algae_invert_swath <- full_algae_invert_swath %>% 
  filter(TYPE %in% c("BARREN", "MACRO")) %>%
  filter(SITE != "SAMSING FOREST")

filtered_algae_invert_swath$TYPE     <- factor(filtered_algae_invert_swath$TYPE,     levels = c("BARREN", "MACRO"))
filtered_algae_invert_swath$SITE     <- factor(filtered_algae_invert_swath$SITE)
filtered_algae_invert_swath$TRANSECT <- factor(filtered_algae_invert_swath$TRANSECT)
filtered_algae_invert_swath$SEGMENT  <- factor(filtered_algae_invert_swath$`SEGMENT`)

# Drop the empty column (Unnamed: 46) and build community matrix
species_cols <- names(filtered_algae_invert_swath)[2:46]  # named indexing is safer than positional

comm <- filtered_algae_invert_swath %>%
  select(all_of(species_cols)) %>%
  mutate(across(everything(), as.numeric))
```

```{r}
names(filtered_algae_invert_swath)
```
