---
title: "soundscapes-data-modeling"
format: html
editor: visual
---

## Forest/Barren SVM

```{r}
all_data$TYPE <- factor(
  all_data$TYPE,
  levels = c("Barren", "Macro_Forest")
)
```

```{r}
# SVM with the same JMP Parameters

library(ggplot2)
library(dplyr)
library(tidyr)
library(kernlab)

BARREN_LABEL <- "Barren"
FOREST_LABEL <- "Macro_Forest"
predictors   <- c("ACI", "NDSI", "AEI", "ADI")
type_col     <- "TYPE"

# Prep data 
keep_rows  <- complete.cases(all_data[, c(predictors, type_col)])
model_data <- all_data[keep_rows, ]
model_data[[type_col]] <- factor(model_data[[type_col]],
                                  levels = c(BARREN_LABEL, FOREST_LABEL))

cat("Total rows used:", nrow(model_data), "\n")
cat("Class counts:\n")
print(table(model_data[[type_col]]))

# Train SVM (cost = C = 1, gamma = sigma = 0.25, RBF kernel)
svm_model <- ksvm(
  TYPE ~ ACI + NDSI + AEI + ADI,
  data       = model_data,
  kernel     = "rbfdot",
  kpar       = list(sigma = 0.25),
  C          = 1,
  prob.model = TRUE,
  scaled     = TRUE
)

cat("\n--- Model Summary ---\n")
cat("Kernel:                   Radial Basis Function\n")
cat("Cost (C):                ", 1, "\n")
cat("Gamma (sigma):           ", 0.25, "\n")
cat("Number of Support Vectors:", svm_model@nSV, "\n")

# 3. Training predictions & confusion matrix 
train_pred <- predict(svm_model, newdata = model_data)
stopifnot(length(train_pred) == nrow(model_data))

conf_mat  <- table(Actual = model_data[[type_col]], Predicted = train_pred)
conf_rate <- prop.table(conf_mat, margin = 1)

cat("\n--- Confusion Matrix (Counts) ---\n")
print(conf_mat)
cat("\n--- Confusion Matrix (Predicted Rates) ---\n")
print(round(conf_rate, 3))

misclass_rate <- 1 - sum(diag(conf_mat)) / sum(conf_mat)
cat("\nMisclassification Rate:", round(misclass_rate, 7), "\n")
cat("(JMP reference:          0.2158539)\n")
```

```{r}
# PLOTS
  # Response Profile Plot (ACI vs NDSI decision boundary)
AEI_fixed <- mean(model_data$AEI, na.rm = TRUE)
ADI_fixed <- mean(model_data$ADI, na.rm = TRUE)

grid <- expand.grid(
  NDSI = seq(min(model_data$NDSI), max(model_data$NDSI), length.out = 150),
  ACI  = seq(min(model_data$ACI),  max(model_data$ACI),  length.out = 150)
) %>%
  mutate(AEI = AEI_fixed, ADI = ADI_fixed)

grid$pred_class <- as.character(predict(svm_model, newdata = grid))

p_profile <- ggplot(grid, aes(x = NDSI, y = ACI, fill = pred_class)) +
  geom_raster(interpolate = TRUE, alpha = 0.75) +
  scale_fill_manual(
    values = c("Barren"       = "#F4CCCC",
               "Macro_Forest" = "#C9DAF8"),
    name   = "TYPE",
    labels = c("Barren"       = "TYPE = Barren",
               "Macro_Forest" = "TYPE = Macro_Forest")
  ) +
  labs(
    title    = "Response Profile Plot",
    subtitle = paste0("AEI fixed at ", round(AEI_fixed, 4),
                      "  |  ADI fixed at ", round(ADI_fixed, 4)),
    x = "NDSI",
    y = "ACI"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "right")

print(p_profile)

# Confusion matrix summary
cat("  Confusion Matrix Summary\n")
cat(sprintf("  %-15s  %-20s  %-20s  %-10s\n",
            "Actual", "Pred:Barren", "Pred:Forest", "Misclass"))
for (lv in c(BARREN_LABEL, FOREST_LABEL)) {
  rc <- conf_mat[lv, ]
  rr <- conf_rate[lv, ]
  cat(sprintf("  %-15s  %-20s  %-20s  %-10s\n",
              lv,
              paste0(rc[BARREN_LABEL], " (", round(rr[BARREN_LABEL], 3), ")"),
              paste0(rc[FOREST_LABEL], " (", round(rr[FOREST_LABEL], 3), ")"),
              round(1 - rr[lv], 3)))
}
cat(sprintf("\n  Overall Misclassification Rate: %.7f\n", misclass_rate))

# Probability Outputs
prob_all <- predict(svm_model, newdata = model_data, type = "probabilities")
model_data$prob_barren <- prob_all[, BARREN_LABEL]
model_data$prob_forest <- prob_all[, FOREST_LABEL]
model_data$gradient_score <- model_data$prob_barren  # 0 = forest, 1 = barren

# Habitat type bar plot
type_summary <- model_data %>%
  group_by(TYPE) %>%
  summarise(
    mean_prob_barren = mean(prob_barren),
    mean_prob_forest = mean(prob_forest),
    .groups = "drop"
  ) %>%
  tidyr::pivot_longer(
    cols = c(mean_prob_barren, mean_prob_forest),
    names_to  = "predicted_as",
    values_to = "mean_prob"
  ) %>%
  mutate(
    predicted_as = if_else(predicted_as == "mean_prob_barren",
                           "Predicted to be urchin barrens",
                           "Predicted to be Macrocystis Forest"),
    TYPE_label = if_else(TYPE == BARREN_LABEL,
                         "Urchin Barrens",
                         "Macrocystis Forest")  
  )

p_jmp <- ggplot(jmp_summary,
                aes(x = TYPE_label, y = mean_prob, fill = predicted_as)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.65) +
  geom_text(aes(label = round(mean_prob, 2)),
            position = position_dodge(width = 0.7),
            vjust = -0.4, size = 4) +
  scale_fill_manual(
    values = c("Predicted to be urchin barrens"     = "#B22222",
               "Predicted to be Macrocystis Forest" = "#4472C4"),
    labels = c("Predicted to be urchin barrens"     = "Predicted to be Urchin Barrens",
               "Predicted to be Macrocystis Forest" = expression("Predicted to be"~italic("Macrocystis")~"Forest")),
    name = NULL
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    expand = expansion(mult = c(0, 0.08)),
    name   = "Probability of assignment (SVM)"
  ) +
  xlab(NULL) +
  theme_bw(base_size = 13) +
  theme(legend.position = "top")

print(p_jmp)

# Site-level mean gradient scores 
gradient_summary <- model_data %>%
  group_by(site, TYPE) %>%
  summarise(
    mean_gradient_score = mean(gradient_score),
    se_gradient_score   = sd(gradient_score) / sqrt(n()),
    .groups = "drop"
  ) %>%
  mutate(
    # Clean site labels: replace _ with space, title case
    site_label = gsub("_", " ", site),
    site_label = tools::toTitleCase(site_label),
    # Clean TYPE labels for legend
    type_label = if_else(TYPE == BARREN_LABEL, "Urchin Barren", "Macrocystis Forest")
  )

# Legend labels with italics for Macrocystis using expression/parse
legend_labels <- c(
  "Urchin Barren"      = "Urchin Barren",
  "Macrocystis Forest" = "italic(Macrocystis)~Forest"
)

p_gradient <- ggplot(gradient_summary,
                     aes(x = reorder(site_label, mean_gradient_score),
                         y = mean_gradient_score,
                         fill = type_label)) +
  geom_col(width = 0.7) +
  geom_errorbar(
    aes(ymin = mean_gradient_score - se_gradient_score,
        ymax = mean_gradient_score + se_gradient_score),
    width = 0.2
  ) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(
    values = c("Urchin Barren"      = "firebrick3",
               "Macrocystis Forest" = "steelblue3"),
    name   = "TYPE",
    labels = c("Urchin Barren"      = "Urchin Barren",
               "Macrocystis Forest" = expression(italic("Macrocystis")~"Forest"))
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    expand = expansion(mult = c(0, 0.05)),
    name   = "Mean gradient score  (0 = forest  \u2192  1 = barren)"
  ) +
  xlab("Site") +
  ggtitle("Mean gradient score by site",
          subtitle = "Error bars = \u00B1 SE; dashed line = classification threshold") +
  theme_bw(base_size = 13) +
  theme(axis.text.x = ggplot2::element_text(angle = 30, hjust = 1))

print(p_gradient)
# Site level summary table
site_summary <- model_data %>%
  group_by(site) %>%
  summarise(
    true_type           = first(TYPE),
    mean_gradient_score = mean(gradient_score),
    n_sequences         = n(),
    .groups = "drop"
  ) %>%
  mutate(
    pred_class = if_else(mean_gradient_score >= 0.5, BARREN_LABEL, FOREST_LABEL),
    correct    = (true_type == pred_class)
  ) %>%
  arrange(mean_gradient_score)

cat("\n--- Site-Level Gradient Summary ---\n")
print(site_summary)

# Save plots
ggsave("svm_barplot.png",      plot = p_jmp,      width = 7, height = 6, dpi = 150)
ggsave("svm_response_profile.png", plot = p_profile,  width = 7, height = 6, dpi = 150)
ggsave("svm_gradient_scores.png",  plot = p_gradient, width = 8, height = 5, dpi = 150)
cat("\nDone. Plots saved.\n")
```

## More Plots!

### Violin Plot of Acoustic Indices by Habitat Type

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)

# Pivot to long format
long_data <- all_data %>%
  select(TYPE, NDSI, ACI, ADI, AEI) %>%
  pivot_longer(cols = c(NDSI, ACI, ADI, AEI),
               names_to = "index",
               values_to = "value")

# Keep index order
long_data$index <- factor(long_data$index, levels = c("NDSI", "ACI", "ADI", "AEI"))

ggplot(long_data, aes(x = TYPE, y = value, fill = index, color = index)) +
  geom_violin(linewidth = 0.8, trim = FALSE) +
  facet_wrap(~ index, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = c(
    "NDSI" = "#648FFF",
    "ACI"  = "#DC267F",
    "ADI"  = "#FE6100",
    "AEI"  = "#F9DE56"
  )) +
  scale_color_manual(values = c(
     "NDSI" = "#648FFF",
    "ACI"  = "#DC267F",
    "ADI"  = "#FE6100",
    "AEI"  = "#F9DE56"
  )) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(x = "Habitat Type", y = NULL)
```

### SVM Probability Plot By Index

```{r}
library(kernlab)
library(ggplot2)
library(tidyr)
library(dplyr)
library(patchwork)

# Extract barren probability
prob_barren <- probs[, "Barren"]

# Build plotting dataframe
plot_df <- model_data %>%
  select(NDSI, ACI, ADI, AEI) %>%
  mutate(prob_barren = prob_barren)

# Function to make one panel
make_panel <- function(df, index, color) {
  ggplot(df, aes(x = .data[[index]], y = prob_barren)) +
    geom_point(color = color, alpha = 0.3, size = 0.8) +
    geom_smooth(color = "black", se = FALSE, linewidth = 0.8) +
    theme_classic() +
    labs(x = index, y = NULL)
}

# Build each panel
p1 <- make_panel(plot_df, "NDSI", "#648FFF")
p2 <- make_panel(plot_df, "ACI",  "#DC267F")
p3 <- make_panel(plot_df, "ADI",  "#FE6100")
p4 <- make_panel(plot_df, "AEI",  "#F9DE56")

# Combine
wrap_plots(p1, p2, p3, p4, nrow = 1) +
  plot_layout(axis_titles = "collect") &
  labs(y = "Probability of assignment as Urchin Barrens (SVM)") &
  theme(plot.margin = margin(5, 5, 5, 5))
```

## Urchin Activity Acoustics Modeling

These acoustics are from three red urchins placed on different substrate/food types (rocks with crustose coralline algae, bare rock, and bare rock with kelp), and recorded underwater (about 30ft depth) for 2 days each.

### CCA SVM

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(kernlab)
library(here)

# Read in data
cca_urchin_data <- read.csv("/Users/aminashenasa/soundscapes_data/data/acoustic_data/filtered_data/Sage_Constraints_Experiment_CCA Urchin Indices.csv")

# Clean TYPE and fix NDSI class
cca_urchin_data$TYPE <- trimws(cca_urchin_data$TYPE)
cca_urchin_data$NDSI <- as.numeric(cca_urchin_data$NDSI)

URCHIN_LABEL    <- "URCHIN_CCA"
NO_URCHIN_LABEL <- "NO URCHIN_CCA"
predictors      <- c("ACI", "NDSI")
type_col        <- "TYPE"

# Prep data
keep_rows             <- complete.cases(cca_urchin_data[, c(predictors, type_col)])
cca_urchin_model_data <- cca_urchin_data[keep_rows, ]
cca_urchin_model_data <- cca_urchin_model_data %>% filter(!is.na(TYPE))
cca_urchin_model_data[[type_col]] <- factor(
  cca_urchin_model_data[[type_col]],
  levels = c(URCHIN_LABEL, NO_URCHIN_LABEL)
)

cat("Total rows used:", nrow(cca_urchin_model_data), "\n")
cat("Class counts:\n")
print(table(cca_urchin_model_data[[type_col]]))

# Build clean dataframe
cca_clean_data <- data.frame(
  TYPE = ifelse(trimws(cca_urchin_data$TYPE) == "URCHIN_CCA", "URCHIN_CCA", "NO_URCHIN_CCA"),
  ACI  = as.numeric(cca_urchin_data$ACI),
  NDSI = as.numeric(cca_urchin_data$NDSI)
)
cca_clean_data <- na.omit(cca_clean_data)
cca_clean_data$TYPE <- factor(
  cca_clean_data$TYPE,
  levels = c("URCHIN_CCA", "NO_URCHIN_CCA")
)

cat("Clean data class counts:\n")
print(table(cca_clean_data$TYPE))
nrow(cca_clean_data)

# Train SVM
cca_urchin_svm_model <- kernlab::ksvm(
  TYPE ~ ACI + NDSI,
  data       = cca_clean_data,
  kernel     = "rbfdot",
  kpar       = list(sigma = 0.25),
  C          = 1,
  prob.model = TRUE,
  scaled     = TRUE
)

cat("\n--- Model Summary ---\n")
cat("Kernel:                   Radial Basis Function\n")
cat("Cost (C):                ", 1, "\n")
cat("Gamma (sigma):           ", 0.25, "\n")
cat("Number of Support Vectors:", cca_urchin_svm_model@nSV, "\n")

# Training predictions & confusion matrix
cca_train_pred <- predict(cca_urchin_svm_model, newdata = cca_clean_data)
cca_conf_mat   <- table(Actual = cca_clean_data$TYPE, Predicted = cca_train_pred)
cca_conf_rate  <- prop.table(cca_conf_mat, margin = 1)

cat("\n--- Confusion Matrix (Counts) ---\n")
print(cca_conf_mat)
cat("\n--- Confusion Matrix (Predicted Rates) ---\n")
print(round(cca_conf_rate, 3))

cca_misclass_rate <- 1 - sum(diag(cca_conf_mat)) / sum(cca_conf_mat)
cat("\nMisclassification Rate:", round(cca_misclass_rate, 7), "\n")

# 5-fold cross validation
set.seed(123)
cca_urchin_svm_cv <- kernlab::ksvm(
  TYPE ~ ACI + NDSI,
  data       = cca_clean_data,
  kernel     = "rbfdot",
  kpar       = list(sigma = 0.25),
  C          = 1,
  cross      = 5,
  scaled     = TRUE
)
cat("5-fold CV error:", cross(cca_urchin_svm_cv), "\n")
cat("5-fold CV accuracy:", 1 - cross(cca_urchin_svm_cv), "\n")
```

```{r}
library(ggplot2)
library(tidyr)

cca_urchin_model_data$NDSI <- as.numeric(cca_urchin_model_data$NDSI)

cca_urchin_model_data %>%
  select(TYPE, ACI, NDSI) %>%
  pivot_longer(cols = c(ACI, NDSI), names_to = "index", values_to = "value") %>%
  ggplot(aes(x = TYPE, y = value, fill = TYPE)) +
  geom_violin() +
  geom_jitter(width = 0.1, alpha = 0.5, size = 2) +
  facet_wrap(~ index, scales = "free_y") +
  scale_fill_manual(values = c("URCHIN_CCA" = "#648FFF", "NO URCHIN_CCA" = "#DC267F")) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(x = NULL, y = "Value")
```

```{r}
# Create prediction grid over ACI and NDSI ranges
ACI_range  <- seq(min(cca_clean_data$ACI), max(cca_clean_data$ACI), length.out = 200)
NDSI_range <- seq(min(cca_clean_data$NDSI), max(cca_clean_data$NDSI), length.out = 200)

cca_grid <- expand.grid(ACI = ACI_range, NDSI = NDSI_range)

cca_grid$pred_class <- predict(cca_urchin_svm_model, newdata = cca_grid)

cca_p_profile <- ggplot(cca_grid, aes(x = NDSI, y = ACI, fill = pred_class)) +
  geom_raster(interpolate = TRUE, alpha = 0.75) +
  geom_point(
    data        = cca_clean_data,
    aes(x = NDSI, y = ACI, shape = TYPE),
    inherit.aes = FALSE,
    size        = 3
  ) +
  scale_fill_manual(
    values = c("URCHIN_CCA"    = "#F4CCCC",
               "NO_URCHIN_CCA" = "#C9DAF8"),
    name   = "Predicted Class",
    labels = c("URCHIN_CCA"    = "Urchins on CCA",
               "NO_URCHIN_CCA" = "Background Noise")
  ) +
  scale_shape_manual(
    values = c("URCHIN_CCA"    = 16,
               "NO_URCHIN_CCA" = 17),
    name   = "Actual Class",
    labels = c("URCHIN_CCA"    = "Urchins on CCA",
               "NO_URCHIN_CCA" = "Background Noise")
  ) +
  labs(
    title = "Response Profile Plot — CCA",
    x     = "NDSI",
    y     = "ACI"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "right")

print(cca_p_profile)
```

### Bare Rock SVM

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(kernlab)
library(here)

# Read in data
rock_urchin_data <- read.csv("/Users/aminashenasa/soundscapes_data/data/acoustic_data/filtered_data/Sage_Constraints_Experiment_Bare Rock Urchin Indices.csv")

# Fix NDSI class
rock_urchin_data$NDSI <- as.numeric(rock_urchin_data$NDSI)

URCHIN_LABEL    <- "URCHINS_BARE ROCK"
NO_URCHIN_LABEL <- "NO URCHINS_BARE ROCK"
predictors      <- c("ACI", "NDSI")
type_col        <- "TYPE"

# Prep data
keep_rows              <- complete.cases(rock_urchin_data[, c(predictors, type_col)])
rock_urchin_model_data  <- urchin_data[keep_rows, ]
rock_urchin_model_data  <- cca_urchin_model_data %>% filter(!is.na(TYPE))
rock_urchin_model_data[[type_col]] <- factor(
  rock_urchin_model_data[[type_col]],
  levels = c(URCHIN_LABEL, NO_URCHIN_LABEL)
)

cat("Total rows used:", nrow(rock_urchin_model_data), "\n")
cat("Class counts:\n")
print(table(rock_urchin_model_data[[type_col]]))

# Start from raw data
rock_clean_data <- data.frame(
  TYPE = gsub(" ", "_", as.character(rock_urchin_data$TYPE)),
  ACI  = as.numeric(rock_urchin_data$ACI),
  NDSI = as.numeric(rock_urchin_data$NDSI)
)

# Remove NAs
rock_clean_data <- na.omit(rock_clean_data)

# Factor with corrected levels
rock_clean_data$TYPE <- factor(
  rock_clean_data$TYPE,
  levels = c("URCHINS_BARE_ROCK", "NO_URCHINS_BARE_ROCK")
)

table(rock_clean_data$TYPE)
nrow(rock_clean_data)
head(rock_clean_data)


# Train SVM
rock_urchin_svm_model <- kernlab::ksvm(
  TYPE ~ ACI + NDSI,
  data       = rock_clean_data,
  kernel     = "rbfdot",
  kpar       = list(sigma = 0.25),
  C          = 1,
  prob.model = TRUE,
  scaled     = TRUE
)

cat("\n--- Model Summary ---\n")
cat("Kernel:                   Radial Basis Function\n")
cat("Cost (C):                ", 1, "\n")
cat("Gamma (sigma):           ", 0.25, "\n")
cat("Number of Support Vectors:", rock_urchin_svm_model@nSV, "\n")

# Training predictions & confusion matrix
rock_train_pred <- predict(rock_urchin_svm_model, newdata = rock_clean_data)
rock_conf_mat   <- table(Actual = rock_clean_data$TYPE, Predicted = rock_train_pred)
rock_conf_rate  <- prop.table(rock_conf_mat, margin = 1)

cat("\n--- Confusion Matrix (Counts) ---\n")
print(rock_conf_mat)
cat("\n--- Confusion Matrix (Predicted Rates) ---\n")
print(round(rock_conf_rate, 3))

rock_misclass_rate <- 1 - sum(diag(rock_conf_mat)) / sum(rock_conf_mat)
cat("\nMisclassification Rate:", round(rock_misclass_rate, 7), "\n")

# 5-fold cross validation
set.seed(123)
rock_urchin_svm_cv <- kernlab::ksvm(
  TYPE ~ ACI + NDSI,
  data       = rock_clean_data,
  kernel     = "rbfdot",
  kpar       = list(sigma = 0.25),
  C          = 1,
  cross      = 5,
  scaled     = TRUE
)
cat("5-fold CV error:", cross(rock_urchin_svm_cv), "\n")
cat("5-fold CV accuracy:", 1 - cross(rock_urchin_svm_cv), "\n")
```

```{r}
# Create prediction grid over ACI and NDSI ranges
ACI_range  <- seq(min(rock_clean_data$ACI), max(rock_clean_data$ACI), length.out = 200)
NDSI_range <- seq(min(rock_clean_data$NDSI), max(rock_clean_data$NDSI), length.out = 200)

rock_grid <- expand.grid(ACI = ACI_range, NDSI = NDSI_range)

# Predict class for each grid point
rock_grid$pred_class <- predict(rock_urchin_svm_model, newdata = rock_grid)

# Plot
rock_p_profile <- ggplot(rock_grid, aes(x = NDSI, y = ACI, fill = pred_class)) +
  geom_raster(interpolate = TRUE, alpha = 0.75) +
  geom_point(
    data        = rock_clean_data,
    aes(x = NDSI, y = ACI, shape = TYPE),
    inherit.aes = FALSE,
    size        = 3
  ) +
  scale_fill_manual(
    values = c("URCHINS_BARE_ROCK"    = "#F4CCCC",
               "NO_URCHINS_BARE_ROCK" = "#C9DAF8"),
    name   = "Predicted Class",
    labels = c("URCHINS_BARE_ROCK"    = "Urchins on bare rock",
               "NO_URCHINS_BARE_ROCK" = "Background noise")
  ) +
  scale_shape_manual(
    values = c("URCHINS_BARE_ROCK"    = 16,
               "NO_URCHINS_BARE_ROCK" = 17),
    name   = "Actual Class",
    labels = c("URCHINS_BARE_ROCK"    = "Urchins on bare rock",
               "NO_URCHINS_BARE_ROCK" = "Background noise")
  ) +
  labs(
    title = "Response Profile Plot — Bare Rock",
    x     = "NDSI",
    y     = "ACI"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "right")

print(rock_p_profile)
```

### Kelp on Bare Rock SVM

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(kernlab)
library(here)

# Read in data
kelp_urchin_data <- read.csv("/Users/aminashenasa/soundscapes_data/data/acoustic_data/filtered_data/Sage_Constraints_Experiment_Kelp Urchin Indices.csv")

# Fix NDSI class
kelp_urchin_data$NDSI <- as.numeric(kelp_urchin_data$NDSI)

URCHIN_LABEL    <- "URCHINS_KELP"
NO_URCHIN_LABEL <- "NO URCHINS_KELP"
predictors      <- c("ACI", "NDSI")
type_col        <- "TYPE"

# Prep data
keep_rows              <- complete.cases(kelp_urchin_data[, c(predictors, type_col)])
kelp_urchin_model_data  <- kelp_urchin_data[keep_rows, ]
kelp_urchin_model_data  <- kelp_urchin_model_data %>% filter(!is.na(TYPE))
kelp_urchin_model_data[[type_col]] <- factor(
  kelp_urchin_model_data[[type_col]],
  levels = c(URCHIN_LABEL, NO_URCHIN_LABEL)
)

cat("Total rows used:", nrow(kelp_urchin_model_data), "\n")
cat("Class counts:\n")
print(table(kelp_urchin_model_data[[type_col]]))

# Start from raw data
kelp_clean_data <- data.frame(
  TYPE = ifelse(kelp_urchin_data$TYPE == "URCHIN_KELP", "URCHIN_KELP", "NO_URCHIN_KELP"),
  ACI  = as.numeric(kelp_urchin_data$ACI),
  NDSI = as.numeric(kelp_urchin_data$NDSI)
)

kelp_clean_data <- na.omit(kelp_clean_data)

kelp_clean_data$TYPE <- factor(
  kelp_clean_data$TYPE,
  levels = c("URCHIN_KELP", "NO_URCHIN_KELP")
)

table(kelp_clean_data$TYPE)
nrow(kelp_clean_data)
head(kelp_clean_data)


# Train SVM
kelp_urchin_svm_model <- kernlab::ksvm(
  TYPE ~ ACI + NDSI,
  data       = kelp_clean_data,
  kernel     = "rbfdot",
  kpar       = list(sigma = 0.25),
  C          = 1,
  prob.model = TRUE,
  scaled     = TRUE
)

cat("\n--- Model Summary ---\n")
cat("Kernel:                   Radial Basis Function\n")
cat("Cost (C):                ", 1, "\n")
cat("Gamma (sigma):           ", 0.25, "\n")
cat("Number of Support Vectors:", kelp_urchin_svm_model@nSV, "\n")

# Training predictions & confusion matrix
kelp_train_pred <- predict(kelp_urchin_svm_model, newdata = kelp_clean_data)
kelp_conf_mat   <- table(Actual = kelp_clean_data$TYPE, Predicted = kelp_train_pred)
kelp_conf_rate  <- prop.table(kelp_conf_mat, margin = 1)

cat("\n--- Confusion Matrix (Counts) ---\n")
print(kelp_conf_mat)
cat("\n--- Confusion Matrix (Predicted Rates) ---\n")
print(round(kelp_conf_rate, 3))

kelp_misclass_rate <- 1 - sum(diag(kelp_conf_mat)) / sum(kelp_conf_mat)
cat("\nMisclassification Rate:", round(kelp_misclass_rate, 7), "\n")

# 5-fold cross validation
set.seed(123)
kelp_urchin_svm_cv <- kernlab::ksvm(
  TYPE ~ ACI + NDSI,
  data       = kelp_clean_data,
  kernel     = "rbfdot",
  kpar       = list(sigma = 0.25),
  C          = 1,
  cross      = 5,
  scaled     = TRUE
)
cat("5-fold CV error:", cross(kelp_urchin_svm_cv), "\n")
cat("5-fold CV accuracy:", 1 - cross(kelp_urchin_svm_cv), "\n")
```

```{r}
# Create prediction grid over ACI and NDSI ranges
ACI_range  <- seq(min(kelp_clean_data$ACI), max(kelp_clean_data$ACI), length.out = 200)
NDSI_range <- seq(min(kelp_clean_data$NDSI), max(kelp_clean_data$NDSI), length.out = 200)

kelp_grid <- expand.grid(ACI = ACI_range, NDSI = NDSI_range)

# Predict class for each grid point
kelp_grid$pred_class <- predict(kelp_urchin_svm_model, newdata = kelp_grid)

# Plot
kelp_p_profile <- ggplot(kelp_grid, aes(x = NDSI, y = ACI, fill = pred_class)) +
  geom_raster(interpolate = TRUE, alpha = 0.75) +
  geom_point(
    data        = kelp_clean_data,
    aes(x = NDSI, y = ACI, shape = TYPE),
    inherit.aes = FALSE,
    size        = 3
  ) +
  scale_fill_manual(
    values = c("URCHIN_KELP"    = "#F4CCCC",
               "NO_URCHIN_KELP" = "#C9DAF8"),
    name   = "Predicted Class",
    labels = c("URCHIN_KELP"    = "Urchins on Kelp",
               "NO_URCHIN_KELP" = "Background Noise")
  ) +
  scale_shape_manual(
    values = c("URCHIN_KELP"    = 16,
               "NO_URCHIN_KELP" = 17),
    name   = "Actual Class",
    labels = c("URCHIN_KELP"    = "Urchins on Kelp",
               "NO_URCHIN_KELP" = "Background Noise")
  ) +
  labs(
    title = "Response Profile Plot — Kelp",
    x     = "NDSI",
    y     = "ACI"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "right")

print(kelp_p_profile)
```

### Principle component with all 3 SVMs

```{r}
library(ggplot2)

# Combine all three datasets with a substrate label
pca_data <- rbind(
  data.frame(cca_clean_data,  substrate = "CCA"),
  data.frame(rock_clean_data, substrate = "Bare Rock"),
  data.frame(kelp_clean_data, substrate = "Kelp")
)

# Run PCA on ACI and NDSI
pca_result <- prcomp(pca_data[, c("ACI", "NDSI")], scale. = TRUE)

summary(pca_result)

# Build plotting dataframe
pca_scores <- as.data.frame(pca_result$x)
pca_scores$substrate <- pca_data$substrate
pca_scores$TYPE      <- pca_data$TYPE

# Plot
ggplot(pca_scores, aes(x = PC1, y = PC2, color = substrate, shape = TYPE)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(aes(group = substrate), level = 0.95, linetype = "dashed") +
  scale_color_manual(values = c(
    "CCA"       = "#E69F00",
    "Bare Rock" = "#0072B2",
    "Kelp"      = "#009E73"
  )) +
  theme_classic() +
  labs(
    title    = "PCA of Urchin Grazing Acoustic Indices",
    x        = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "% variance)"),
    y        = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "% variance)"),
    color    = "Substrate",
    shape    = "Class"
  ) +
  theme(legend.position = "right")
head(kelp_probs)
```
