---
title: "ecological-data-modeling"
format: html
editor: visual
---

## Ecological Data Modeling

### PERMANOVA

```{r}
library(vegan)
library(dplyr)
library(readxl)
library(ggplot2)

# PERMANOVA
set.seed(123)

permanova <- adonis2(
  comm ~ TYPE + SITE %in% TYPE + TRANSECT %in% SITE + SEGMENT %in% TRANSECT,
  data         = filtered_algae_invert_swath,
  method       = "bray",
  permutations = 999
)

print(permanova)
```

```{r}
# Homogeneity of dispersion
dist_bm <- vegdist(comm, method = "bray")

dispersion <- betadisper(dist_bm, filtered_algae_invert_swath$TYPE)
permutest(dispersion)
plot(dispersion)
```

### NMDS

```{r}
library(ggtext)
hab_cols_named <- c("BARREN" = "#F8766D", "MACRO" = "#00BFC4")

community_nmds <- ggplot(scores_df, aes(x = NMDS1, y = NMDS2, color = TYPE, shape = SITE)) +
  geom_point(size = 3, alpha = 1) +
  scale_shape_manual(
    values = c(
      "BATTERY ISLAND" = 3,  "CROW ISLAND"  = 7,  "MIDDLE ISLAND" = 8,
      "HARRIS ISLAND"  = 16, "SAMSING COVE" = 17, "SANDY COVE"    = 15
    ),
    labels = site_labels,
    limits = c(
      "BATTERY ISLAND", "CROW ISLAND", "MIDDLE ISLAND",
      "HARRIS ISLAND", "SAMSING COVE", "SANDY COVE"
    )
  ) +
  stat_ellipse(aes(group = TYPE), level = 0.95, linetype = "dashed") +
  scale_color_manual(
    values = hab_cols_named,
    labels = c("BARREN" = "Urchin Barren", "MACRO" = "*Macrocystis* Forest")
  ) +
  theme_urchin(base_size = 13) +
  theme(
    legend.position     = "right",
    panel.grid.major.y  = element_blank(),   # remove horizontal gridlines
    panel.grid.major.x  = element_blank(),   # already off in theme, kept explicit for clarity
    axis.line           = element_line(colour = "grey20", linewidth = 0.4)
  ) +
  labs(
    title = paste0("NMDS of community structure (stress = ",
                   round(nmds$stress, 3), ")"),
    color = "Habitat Type",
    shape = "Site"
  )

community_nmds

ggsave("community_nmds.pdf", community_nmds)
```

```{r}
# Extract top contributors directly from the simper object
simper_summary <- summary(simper_result)  # simper_result = your simper() output
simper_top <- simper_summary$BARREN_MACRO  # replace with your actual comparison name, e.g. summary(simper_result)$"BARREN_MACRO"

# Keep top N contributors by cumulative contribution or raw average, filtered to significant ones
simper_species <- simper_top |>
  tibble::rownames_to_column("species") |>
  dplyr::filter(p <= 0.05) |>
  dplyr::slice_max(average, n = 5)

# Match these species names to their NMDS species scores (from your ordination)
species_scores_top <- as.data.frame(scores(nmds, display = "species"))
species_scores_top$species <- rownames(species_scores_top)
species_scores_top <- dplyr::semi_join(species_scores_top, simper_species, by = "species")
```

```{r}
community_nmds_vec <- ggplot(scores_df, aes(x = NMDS1, y = NMDS2, color = TYPE, shape = SITE)) +
  geom_point(size = 3, alpha = 1) +
  scale_shape_manual(
    values = c(
      "BATTERY ISLAND" = 3,  "CROW ISLAND"  = 7,  "MIDDLE ISLAND" = 8,
      "HARRIS ISLAND"  = 16, "SAMSING COVE" = 17, "SANDY COVE"    = 15
    ),
    labels = site_labels,
    limits = c(
      "BATTERY ISLAND", "CROW ISLAND", "MIDDLE ISLAND",
      "HARRIS ISLAND", "SAMSING COVE", "SANDY COVE"
    )
  ) +
  stat_ellipse(aes(group = TYPE), level = 0.95, linetype = "dashed") +
  geom_segment(
    data = species_scores_top,
    aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2),
    inherit.aes = FALSE,
    arrow = arrow(length = unit(0.2, "cm")),
    color = "grey30", linewidth = 0.15
  ) +
  scale_color_manual(
    values = hab_cols_named,
    labels = c("BARREN" = "Urchin Barren", "MACRO" = "*Macrocystis* Forest")
  ) +
  labs(
    title = paste0("NMDS of community structure (stress = ",
                   round(nmds$stress, 3), ")"),
    color = "Habitat Type",
    shape = "Site"
  ) +
  theme_urchin(base_size = 13) +
  theme(
    legend.position    = "right",
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.background   = element_rect(fill = "white", color = NA),
    axis.line          = element_line(colour = "grey20", linewidth = 0.4)
  )

community_nmds_vec

ggsave("community_nmds_vectors.pdf", community_nmds_vec)
```

```{r}
community_nmds_vec <- ggplot(scores_df, aes(x = NMDS1, y = NMDS2, color = TYPE, shape = SITE)) +
  geom_point(size = 3, alpha = 1) +
  scale_shape_manual(
    values = c(
      "BATTERY ISLAND" = 3,  "CROW ISLAND"  = 7,  "MIDDLE ISLAND" = 8,
      "HARRIS ISLAND"  = 16, "SAMSING COVE" = 17, "SANDY COVE"    = 15
    ),
    labels = site_labels,
    limits = c(
      "BATTERY ISLAND", "CROW ISLAND", "MIDDLE ISLAND",
      "HARRIS ISLAND", "SAMSING COVE", "SANDY COVE"
    )
  ) +
  stat_ellipse(aes(group = TYPE), level = 0.95, linetype = "dashed") +
  geom_segment(
    data = species_scores_top,
    aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2),
    inherit.aes = FALSE,
    arrow = arrow(length = unit(0.2, "cm")),
    color = "grey30", linewidth = 0.2
  ) +
  geom_text_repel(
    data = species_scores_top,
    aes(x = NMDS1, y = NMDS2, label = species),
    inherit.aes = FALSE,
    color = "grey20", size = 3.2, fontface = "italic",
    segment.color = "grey50", segment.size = 0.2,
    min.segment.length = 0, box.padding = 0.3, point.padding = 0.2,
    max.overlaps = Inf, seed = 42
  ) +
  scale_color_manual(
    values = hab_cols_named,
    labels = c("BARREN" = "Urchin Barren", "MACRO" = "*Macrocystis* Forest")
  ) +
  labs(
    title = paste0("NMDS of community structure (stress = ",
                   round(nmds$stress, 3), ")"),
    color = "Habitat Type",
    shape = "Site"
  ) +
  theme_urchin(base_size = 13) +
  theme(
    legend.position    = "right",
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.background   = element_rect(fill = "white", color = NA),
    axis.line          = element_line(colour = "grey20", linewidth = 0.4)
  )

community_nmds_vec

```

### SIMPER

```{r}
simper_result <- simper(
  comm,
  group        = filtered_algae_invert_swath$TYPE,
  permutations = 999
)

summary(simper_result)
# Top contributors to BARREN vs MACRO dissimilarity
```

### SVM Dissimilarity

```{r}
library(vegan)
library(dplyr)
library(ggplot2)


# Site-level community matrix (aggregate the swath data to 6 sites)
#    comm is per-segment; collapse to one row per SITE for the 6-site analysis.

site_key <- filtered_algae_invert_swath$SITE
comm_site <- rowsum(comm, group = site_key)          # sum abundances per site
# (use aggregate/means instead if your Bray-Curtis was built on mean abundance)

site_names <- rownames(comm_site)

# Community dissimilarity (Bray-Curtis) among the 6 sites
comm_diss <- vegdist(comm_site, method = "bray")


# Acoustic dissimilarity (Euclidean) among the same 6 sites,
#    from the SVM mean barren probability per site.
#    Build a named vector: mean SVM barren prob per site.

prob_by_site_svm <- habitat_model_data %>%
  mutate(prob = prob_barren) %>%
  group_by(site) %>%
  summarise(mean_prob = mean(prob, na.rm = TRUE), .groups = "drop")

# align the acoustic sites to the SAME order as comm_site's rownames
# (normalize case/spacing so the two site labelings match)
norm <- function(x) toupper(trimws(gsub("[_]+", " ", x)))
prob_vec <- prob_by_site_svm$mean_prob[
  match(norm(site_names), norm(prob_by_site_svm$site))
]
names(prob_vec) <- site_names
stopifnot(!any(is.na(prob_vec)))       # halts if any site failed to match

acoustic_diss <- dist(prob_vec, method = "euclidean")   # 15 pairwise distances

# Pair the two dissimilarity sets into 15 rows

pair_labels <- t(combn(site_names, 2))
diss_df <- data.frame(
  site_a    = pair_labels[, 1],
  site_b    = pair_labels[, 2],
  comm_diss = as.vector(comm_diss) * 100,   # scale to 0-100 like the JMP y-axis
  acou_diss = as.vector(acoustic_diss)
)

# label each pair by type combination (MacroMacro / BarrensBarrens / BarrensMacro)
site_type <- filtered_algae_invert_swath %>%
  distinct(SITE, TYPE) %>%
  mutate(SITE = norm(SITE))
type_of <- setNames(site_type$TYPE, site_type$SITE)
lab_pair <- function(a, b) {
  ta <- type_of[norm(a)]; tb <- type_of[norm(b)]
  paste0(sort(c(ifelse(ta=="BARREN","Barrens","Macro"),
                ifelse(tb=="BARREN","Barrens","Macro"))), collapse = "")
}
diss_df$pair_type <- mapply(lab_pair, diss_df$site_a, diss_df$site_b)

# log fit
fit <- lm(comm_diss ~ log(acou_diss), data = diss_df)
s <- summary(fit); f <- s$fstatistic
cat(sprintf("R2 = %.4f  F(%d,%d) = %.2f  p = %.4f\n",
            s$r.squared, f[2], f[3], f[1],
            pf(f[1], f[2], f[3], lower.tail = FALSE)))
cat(sprintf("Community diss = %.4f + %.4f * log(Acoustic diss)\n",
            coef(fit)[1], coef(fit)[2]))

# plot
p_fig8 <- ggplot(diss_df, aes(acou_diss, comm_diss)) +
  geom_smooth(method = "lm", formula = y ~ log(x), se = FALSE,
              colour = "sienna3", linewidth = 1) +
  geom_text(aes(label = pair_type), size = 3, colour = "grey20") +
  labs(x = "Acoustic dissimilarity",
       y = "Community dissimilarity") +
  theme_bw(base_size = 13) +
  theme(panel.grid.minor = element_blank())

print(p_fig8)
ggsave("figure8_svm_dissimilarity.png", p_fig8, width = 8, height = 6, dpi = 150)
```

```{r}
p_fig8 <- ggplot(diss_df, aes(acou_diss, comm_diss)) +
  geom_smooth(method = "lm", formula = y ~ log(x), se = TRUE,
              colour = "black", fill = "grey85", linewidth = 1) +
  geom_point(aes(colour = pair_type), size = 3) +
  scale_colour_manual(
    values = c("MacroMacro"     = "steelblue3",
               "BarrensBarrens" = "firebrick3",
               "BarrensMacro"   = "orchid3"),
    name = "Site pair",
    labels = c("MacroMacro" = "Forest–Forest",
               "BarrensBarrens" = "Barren–Barren",
               "BarrensMacro" = "Forest–Barren")) +
  annotate("text",
           x = -Inf, y = Inf,
           label = lab_text, hjust = -0.1, vjust = 1.5, size = 4, colour = "grey20") +
  labs(x = "Acoustic dissimilarity", y = "Community dissimilarity") +
  theme_bw(base_size = 13) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom")
p_fig8

ggsave("svm_dissimilarity.png", p_fig8,
       width = 9, height = 5.5, dpi = 300)
```

## Ecological plots

```{r}
eco_data <- read.csv(PATH_ECO) %>%
  rename(
    red_urchin_biomass   = `Mean.Mean.Red.Urchin.Biomass...m2.`,
    stipe_density        = `Mean.Mean.Stipe.Density..m.2..`,
    red_urchin_density   = `Mean.Strongylocentrotus.franciscanus..red.urchin.m.2.`,
    green_urchin_density = `Mean.Strongylocentrotus..green.urchin.m.2.`
  ) %>%
  mutate(join_key = tolower(SITE))

ECO_VARS <- c(red_urchin_density   = "Red urchin density (m\u00B2)",
              red_urchin_biomass   = "Red urchin biomass (m\u00B2)",
              green_urchin_density = "Green urchin density (m\u00B2)",
              stipe_density        = "Macrocystis stipe density (m\u00B2)")
```

```{r}
library(dplyr)
# summaries
# Ensure factors and string assignments match perfectly
eco_by_type <- site_data %>%
  mutate(type_label = factor(
    if_else(true_type == "Macro_Forest", "Macrocystis Forest", "Urchin Barren"), # Fix: Capital F
    levels = c("Macrocystis Forest", "Urchin Barren"))) %>%
  group_by(type_label) %>%
  summarise(
    red_mean   = mean(red_urchin_density, na.rm = TRUE),
    red_se     = sd(red_urchin_density,   na.rm = TRUE) / sqrt(n()),
    stipe_mean = mean(stipe_density,      na.rm = TRUE),
    stipe_se   = sd(stipe_density,        na.rm = TRUE) / sqrt(n()),
    .groups = "drop")

# fixed left-to-right site order: forests first, then barrens
SITE_ORDER <- c("Battery Island", "Crow Island", "Middle Island",   # forests
                "Harris Island", "Sandy Cove", "Samsing Cove")       # barrens

# by SITE (one value per site)
eco_by_site <- site_data %>%
  mutate(
    site_label = tools::toTitleCase(gsub("_", " ", site)),
    type_label = factor(
      if_else(true_type == "Macro_Forest", "Macrocystis Forest", "Urchin Barren"),
      levels = c("Macrocystis Forest", "Urchin Barren")),
    site_label = factor(site_label, levels = SITE_ORDER)
  )

# guard: catch any label that doesn't match SITE_ORDER (would silently vanish)
stopifnot(!anyNA(eco_by_site$site_label)) 

```

```{r}
library(ggplot2)
library(ggtext)

# 1. Update your type plots with HTML italics and element_markdown()
p_stipe_type <- ggplot(eco_by_type, aes(type_label, stipe_mean, fill = type_label)) +
  geom_col(width = 0.65) +
  geom_errorbar(aes(ymin = stipe_mean - stipe_se, ymax = stipe_mean + stipe_se),
                width = 0.15, linewidth = 0.4) +
  geom_label(aes(label = sprintf("%.2f", stipe_mean), y = pmax(stipe_mean, 0.06)),
             size = 3.3, colour = "grey20", fill = "white",
             label.size = 0, label.padding = unit(0.1, "lines")) +
  
  # SWAP expressions out for clean HTML styling strings:
  scale_x_discrete(labels = c(
    "Macrocystis Forest" = "<i>Macrocystis</i> forest",
    "Urchin Barren"      = "Urchin barren"
  )) +
  
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(x = NULL, y = expression(italic("Macrocystis") ~ "stipe density (m"^-2*")")) +
  hab_fill() + guides(fill = "none") +
  theme_urchin(base_size = 13) +
  
  # FORCE the theme to allow markdown styling on the x-axis:
  theme(axis.text.x = element_markdown()) 


p_red_type <- ggplot(eco_by_type, aes(type_label, red_mean, fill = type_label)) +
  geom_col(width = 0.65) +
  geom_errorbar(aes(ymin = red_mean - red_se, ymax = red_mean + red_se),
                width = 0.15, linewidth = 0.4) +
  geom_label(aes(label = sprintf("%.2f", red_mean), y = pmax(red_mean, 0.06)),
             size = 3.3, colour = "grey20", fill = "white",
             label.size = 0, label.padding = unit(0.1, "lines")) +
  
  # SWAP expressions out for clean HTML styling strings:
  scale_x_discrete(labels = c(
    "Macrocystis Forest" = "<i>Macrocystis</i> forest",
    "Urchin Barren"      = "Urchin barren"
  )) +
  
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(x = NULL, y = expression("Red urchin density (m"^-2*")")) +
  hab_fill() + guides(fill = "none") +
  theme_urchin(base_size = 13) +
  
  # FORCE the theme to allow markdown styling on the x-axis:
  theme(axis.text.x = element_markdown())

# by site (fixed order)
p_red_site <- ggplot(eco_by_site, aes(site_label, red_urchin_density, fill = type_label)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.2f", red_urchin_density)),
            vjust = -0.5, size = 3, colour = "grey40") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(x = "Site", y = expression("Red urchin density (m"^-2*")")) +
  hab_fill("Site type") +
  theme_urchin(base_size = 13, angle_x = 30)

p_stipe_site <- ggplot(eco_by_site, aes(site_label, stipe_density, fill = type_label)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.2f", stipe_density)),
            vjust = -0.5, size = 3, colour = "grey40") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(x = "Site", y = expression(italic("Macrocystis") ~ "stipe density (m"^-2*")")) +
  hab_fill("Site type") +
  theme_urchin(base_size = 13, angle_x = 30)

p_stipe_type; p_red_type; p_red_site; p_stipe_site

```

```{r}
library(grid)

# Helper function to cleanly export an egg-aligned gtable object to an exact-fit PDF
save_fixed_gtable <- function(filename, gtable_obj) {
  # 1. Open the PDF engine device
  # (No need to provide hardcoded inches; grid automatically reads the gtable layout sizes)
  pdf(filename, 
      width = convertWidth(sum(gtable_obj$widths), "in", valueOnly = TRUE), 
      height = convertHeight(sum(gtable_obj$heights), "in", valueOnly = TRUE))
  
  # 2. Draw the plot cleanly on the native canvas grid
  grid.draw(gtable_obj)
  
  # 3. Safely turn off and write the file
  dev.off()
}

# ---- EXECUTE PERFECTLY-SIZED SAVES ----
save_fixed_gtable("fig_red_type.pdf",   p_red_type_fx)
save_fixed_gtable("fig_stipe_type.pdf", p_stipe_type_fx)
save_fixed_gtable("fig_red_site.pdf",   p_red_site_fx)
save_fixed_gtable("fig_stipe_site.pdf", p_stipe_site_fx)

```

## Map of Sites

Loading packages

```{r}
library(sf)
library(terra)
library(ggplot2)
library(ggspatial)
library(cowplot)
library(maptiles)
library(ggnewscale)   
library(ggstar)
library(ggrepel)
library(rnaturalearth)
library(osmdata)
```

Site data

```{r}
sites <- data.frame(
  site = c("Crow Island","Middle Island","Battery Island",
           "Harris Island","Samsing Cove","Sandy Cove"),
  type = c("Macrocystis forest","Macrocystis forest","Macrocystis forest",
           "Urchin barren","Urchin barren","Urchin barren"),
  x = c(-135.47399,-135.45316,-135.38126,-135.27785,-135.35305,-135.32650),
  y = c( 57.11711,  57.08433,  57.06040,  57.03242,  56.98664,  56.98737)
)

# Sitka Sound sits in UTM zone 8N (EPSG:32608). Working in a projected CRS
# (metres) gives an accurate scale bar and a meaningful north arrow.
utm_crs   <- 32608
sites_sf  <- st_as_sf(sites, coords = c("x","y"), crs = 4326)
sites_utm <- st_transform(sites_sf, utm_crs)
```

```{r}
hab_star <- c("Macrocystis forest" = 15,   # filled circle
              "Urchin barren"      = 1)    # 5-point star
```

```{r}
library(sf)
library(osmdata)
library(ggplot2)
library(ggstar) # Assumed for geom_star
library(cowplot)
library(rnaturalearth)

# ---- 1. SETUP SITES DATA ----
sites <- data.frame(
  site = c("Crow Island","Middle Island","Battery Island",
           "Harris Island","Samsing Cove","Sandy Cove"),
  type = c("Macrocystis forest","Macrocystis forest","Macrocystis forest",
           "Urchin barren","Urchin barren","Urchin barren"),
  x = c(-135.47399,-135.45316,-135.38126,-135.27785,-135.35305,-135.32650),
  y = c( 57.11711,  57.08433,  57.06040,  57.03242,  56.98664,  56.98737)
)

utm_crs   <- 32608
sites_sf  <- st_as_sf(sites, coords = c("x","y"), crs = 4326)
sites_utm <- st_transform(sites_sf, utm_crs)

# ---- 2. DEFINE BOUNDING BOX FRAME ----
view_ratio     <- 16 / 9
view_height_km <- 18

ctr    <- st_coordinates(st_centroid(st_combine(sites_utm)))[1, ]
half_h <- view_height_km * 1000 / 2
half_w <- half_h * view_ratio
frame  <- st_bbox(c(xmin = ctr[["X"]] - half_w, xmax = ctr[["X"]] + half_w,
                    ymin = ctr[["Y"]] - half_h, ymax = ctr[["Y"]] + half_h),
                  crs = utm_crs)

bbox_ll <- st_bbox(st_transform(st_as_sfc(frame), 4326))
pad     <- 0.15
osm_bbox <- c(bbox_ll["xmin"] - pad, bbox_ll["ymin"] - pad,
              bbox_ll["xmax"] + pad, bbox_ll["ymax"] + pad)
names(osm_bbox) <- c("xmin","ymin","xmax","ymax")

options(tigris_class = "sf")
options(tigris_use_cache = TRUE) # Enabled local caching as recommended

# ---- 3. HIGH-RESOLUTION LOCAL TIGER BASEMAP ----
view_ratio     <- 16 / 9
view_height_km <- 18

ctr    <- st_coordinates(st_centroid(st_combine(sites_utm)))[1, ]
half_h <- view_height_km * 1000 / 2
half_w <- half_h * view_ratio
frame  <- st_bbox(c(xmin = ctr[["X"]] - half_w, xmax = ctr[["X"]] + half_w,
                    ymin = ctr[["Y"]] - half_h, ymax = ctr[["Y"]] + half_h),
                  crs = utm_crs)

frame_sfc <- st_as_sfc(frame)

# 1. Download official high-res shapes
alaska_land <- area_water(state = "AK", county = "Sitka", year = 2023)

# 2. Invert and force into a proper full sf data frame layer
coast_polys_filled <- tryCatch({
  water_projected <- st_transform(alaska_land, utm_crs)
  geom_diff <- st_difference(frame_sfc, st_union(water_projected))
  
  # FIX: Convert the raw geometry (sfc) into a true sf object layer
  st_sf(geometry = geom_diff)
}, error = function(e) {
  fallback <- states(year = 2023) |> 
    subset(NAME == "Alaska") |> 
    st_transform(utm_crs) |> 
    st_intersection(frame_sfc)
  st_sf(geometry = st_geometry(fallback))
})

# This assignment now safely succeeds because coast_polys_filled is a true sf object
st_agr(coast_polys_filled) <- "constant"


# ---- 4. AESTHETICS AND MAPPING LAYERS ----
# Map your types directly to the aesthetic settings
hab_star <- c("Macrocystis forest" = 15, "Urchin barren" = 1) 
hab_cols <- c("Macrocystis forest" = "#00BFC4", "Urchin barren" = "#F8766D") 
hab_labs <- c("Macrocystis forest" = expression(italic("Macrocystis") ~ "forest"), "Urchin barren" = "Urchin barren")

# ---- MAIN MAP LAYOUT ----
main <- ggplot() +
  # Draw land polygons
  geom_sf(data = coast_polys_filled, fill = "grey80", color = "grey40", linewidth = 0.3) + 

  # Draw data layers using your loaded sites_utm spatial object
  geom_star(data = sites_utm, aes(geometry = geometry, starshape = type, fill = type),
            stat = "sf_coordinates", size = 4, color = "black", starstroke = 0.5) +

  scale_fill_manual(name = NULL, values = hab_cols, labels = hab_labs) +
  scale_starshape_manual(name = NULL, values = hab_star, labels = hab_labs) +

         # 1. SCALE BAR (Stays cleanly anchored in the bottom right corner)
  ggspatial::annotation_scale(
    location = "br", 
    width_hint = 0.20, 
    pad_x = unit(0.5, "cm"), 
    pad_y = unit(0.4, "cm"),
    text_col = "black", 
    line_col = "black", 
    bar_cols = c("black", "white"),
    style = "ticks", 
    text_cex = 0.8
  ) +
  
  # 2. MOVED NORTH ARROW (Relocated to the bottom-left corner)
  ggspatial::annotation_north_arrow(
    location = "bl", # Changed to Bottom Left
    which_north = "grid",
    height = unit(0.6, "cm"), # Maintained the clean, minimal size
    width = unit(0.6, "cm"),  
    pad_x = unit(0.6, "cm"),  # Slightly pushed away from the left border line
    pad_y = unit(0.4, "cm"),  # Matches the bottom border spacing baseline
    style = ggspatial::north_arrow_orienteering(
      text_col = "black", 
      line_col = "black",
      fill = c("white", "black")
    )
  ) +

  coord_sf(crs = utm_crs,
           xlim = c(frame[["xmin"]], frame[["xmax"]]),
           ylim = c(frame[["ymin"]], frame[["ymax"]]),
           expand = FALSE) +
  theme_minimal(base_size = 11) +
   theme(
    panel.background = element_rect(fill = "white", color = NA), 
    panel.border      = element_rect(color = "black", fill = NA, linewidth = 0.8),
    panel.grid.major  = element_line(color = "grey70", linewidth = 0.2),
    panel.grid.minor  = element_blank(),
    axis.text         = element_text(size = 8, color = "black"),
    axis.title        = element_blank(),
    legend.position   = c(0.85, 0.92),
    legend.title      = element_blank(),  
    legend.text       = element_text(size = 11),
    legend.key        = element_blank(),
    legend.background = element_rect(fill = scales::alpha("white", 0.9), color = NA),
    legend.margin     = margin(4, 8, 4, 8),
    plot.margin       = margin(2, 2, 2, 2)
  )


```

```{r}

# ---- 1. RESTORE ORIGINAL INSET VISUAL SCALING (140 KM BOUNDS) ----
inset_half_km <- 140 
inset_half_w  <- inset_half_km * 1000
inset_half_h  <- inset_half_w

inset_frame <- st_bbox(c(xmin = ctr[["X"]] - inset_half_w, xmax = ctr[["X"]] + inset_half_w,
                          ymin = ctr[["Y"]] - inset_half_h, ymax = ctr[["Y"]] + inset_half_h),
                        crs = utm_crs)

inset_bbox_ll <- st_bbox(st_transform(st_as_sfc(inset_frame), 4326))


# ---- 1. MOVE CAMERA WINDOW WESTWARD (CREATES WHITE SPACE ON THE LEFT) ----
inset_half_km <- 140 
inset_half_w  <- inset_half_km * 1000
inset_half_h  <- inset_half_w

# Subtracted 25,000 meters from the X center to pull more ocean into the left side
inset_frame <- st_bbox(c(xmin = (ctr[["X"]] - 25000) - inset_half_w, 
                          xmax = (ctr[["X"]] - 25000) + inset_half_w,
                          ymin = ctr[["Y"]] - inset_half_h, 
                          ymax = ctr[["Y"]] + inset_half_h),
                        crs = utm_crs)

inset_bbox_ll <- st_bbox(st_transform(st_as_sfc(inset_frame), 4326))


# ---- REPLACES YOUR ENTIRE INSET BLOCK WITH FINAL ALIGNMENTS ----
inset <- ggplot() +
  geom_sf(data = alaska_inset_map, fill = "grey80", color = "grey40", linewidth = 0.2) +
  geom_sf(data = study_box, fill = NA, color = "red", linewidth = 1.0) +
  
  # Southeast Alaska: Scooted left into the open white space to ensure high legibility
  annotate(geom = "text", x = -136.9, y = 57.45, label = "Southeast\nAlaska", 
           size = 3.6, fontface = "italic", color = "black", lineheight = 0.85, 
           vjust = 0) +
  
  # Sitka: Left exactly where it belongs under the red box
    # Sitka: Made bold to stand out clearly from the background map elements
  annotate(geom = "text", x = -135.25, y = 56.80, label = "Sitka", 
           size = 3.6, fontface = "bold", color = "black", 
           hjust = 0, vjust = 0.5) +
  
  # Gulf of Alaska: Cleanly settled in the lower left marine zone
  annotate(geom = "text", x = -137, y = 56.55, label = "Gulf of\nAlaska", 
           size = 3.6, fontface = "italic", color = "grey30", lineheight = 0.85, 
           hjust = 0.5) +
           
  coord_sf(xlim = c(inset_bbox_ll["xmin"], inset_bbox_ll["xmax"]),
           ylim = c(inset_bbox_ll["ymin"], inset_bbox_ll["ymax"]),
           expand = FALSE, datum = NA) +
  theme_void() +
  theme(
    panel.border     = element_rect(color = "black", fill = NA, linewidth = 0.8),
    panel.background = element_rect(fill = "white", color = NA),     
    plot.background  = element_rect(fill = "transparent", color = NA) 
  )


# ---- 2. MAP COMPOSITE LAYOUT (LARGER VISUAL FOOTPRINT) ----
map_composite <- ggdraw() +
  draw_plot(main) +
  # Expanded width/height to make the inset stand out and look crisp
  draw_plot(inset, x = 0.05, y = 0.650, width = 0.26, height = 0.32)

# ---- 3. FINAL PLOT COMBINATION ----
final_output <- plot_grid(
  title_theme,
  map_composite,
  ncol = 1,
  rel_heights = c(0.08, 0.92)
)

# ---- 4. SAVE HIGH-RES OUTPUT ----
out_height <- 8
ggsave("sitka_sound_map.png",
       final_output,
       width = out_height * view_ratio,
       height = out_height,
       dpi = 400,
       bg = "white",
       scale = .6)

# Display final result
final_output

```

```{r}

# Directly assign the composite map as the final output (removes title completely)
final_output <- map_composite

# save high res output
out_height <- 8
ggsave("sitka_sound_map.png",
       final_output,
       width = out_height * view_ratio,
       height = out_height,
       dpi = 400,
       bg = "white",
       scale = .6)

# Display final result in your viewer window
final_output

```

```{r}
# ---- 5. SAVING ----
out_height <- 8
ggsave("sitka_sound_map.png",
       final_output,
       width = out_height * view_ratio,
       height = out_height,
       dpi = 400,
       bg = "white",
       scale = .6)

final_output
```

```{r}
# This prints the exact width of your map canvas in kilometers
as.numeric(frame[["xmax"]] - frame[["xmin"]]) / 1000

```

## Urchin Biomass Equation

To determine red sea urchin biomass, I created a biomass equation from Southeast Alaska red sea urchins to convert sea urchin density values to a more ecologically meaningful measure of grazer impact. Mesocentrotus franciscanus (red sea urchins) were collected from urchin barren sites (Harris Island, Sandy Cove, and Samsing Cove) and from one of the Macrocystis forest sites (Middle Island). Approximately 30 red sea urchins were collected from each urchin barren site (n = 88 total), and 12 from the Macrocystis forest site. Measurements of volume (ml), total body weight (g), test width (mm), and gonad weight (g), were taken from each urchin.

```{r}
library(here)
urchin_gsi <- read.csv("/users/aminashenasa/soundscapes_data/data/ecological_data/urchin_gsi.csv")
```

```{r}
library(ggplot2)

set.seed(42)

urchin_gsi <- urchin_gsi[!is.na(urchin_gsi$Test.Width..mm.) & !is.na(urchin_gsi$Total.Body.Weight..g.), ]

x <- urchin_gsi$Test.Width..mm.
y <- urchin_gsi$Total.Body.Weight..g.

# Exponential 3P model
# y = Asymptote + Scale * exp(GrowthRate * x)
fit <- nls(y ~ Asymptote + Scale * exp(GrowthRate * x),
           start = list(Asymptote = -30, Scale = 20, GrowthRate = 0.03),
           control = nls.control(maxiter = 200))

coefs <- coef(fit)
rss <- sum(residuals(fit)^2)
tss <- sum((y - mean(y))^2)
rsq <- 1 - rss / tss

cat("Coefficients:\n"); print(coefs)
cat("R-squared:", rsq, "\n")

pred_fun <- function(pars, xv) pars["Asymptote"] + pars["Scale"] * exp(pars["GrowthRate"] * xv)

xseq <- seq(min(x), max(x), length.out = 200)
fit_y <- pred_fun(coefs, xseq)

# Bootstrap confidence band (nonparametric residual resampling) 
resid0 <- residuals(fit)
fitted0 <- fitted(fit)
n <- length(y)
B <- 1000
boot_preds <- matrix(NA, nrow = B, ncol = length(xseq))

for (b in seq_len(B)) {
  y_star <- fitted0 + sample(resid0, n, replace = TRUE)
  fit_b <- tryCatch(
    nls(y_star ~ Asymptote + Scale * exp(GrowthRate * x),
        start = as.list(coefs),
        control = nls.control(maxiter = 100, warnOnly = TRUE)),
    error = function(e) NULL
  )
  if (!is.null(fit_b)) {
    boot_preds[b, ] <- pred_fun(coef(fit_b), xseq)
  }
}

ci_lower <- apply(boot_preds, 2, quantile, probs = 0.025, na.rm = TRUE)
ci_upper <- apply(boot_preds, 2, quantile, probs = 0.975, na.rm = TRUE)

pred_df <- data.frame(x = xseq, y = fit_y, lower = ci_lower, upper = ci_upper)

#  Annotation text 
ann_text <- sprintf(
  "y = %.2f + %.2f*exp(%.4f*x)\n R\u00b2 = %.4f",
  coefs["Asymptote"], coefs["Scale"], coefs["GrowthRate"], rsq
)

# Plot
blue <- "#4472C4"
band <- "#B7C9EC"

urchin_biomass <- ggplot() +
  geom_ribbon(data = pred_df, aes(x = x, ymin = lower, ymax = upper),
              fill = jmp_band, alpha = 0.55) +
  geom_line(data = pred_df, aes(x = x, y = y),
            color = jmp_blue, linewidth = 1) +
  geom_point(data = urchin_gsi, aes(x = Test.Width..mm., y = Total.Body.Weight..g.),
             shape = 21, fill = "black", color = "black", size = 2.4, stroke = 0.2) +
annotate("text", x = -Inf, y = Inf, label = ann_text,
         hjust = -0.05, vjust = 1.1, size = 3.6, family = "sans", color = "grey20") +
  scale_x_continuous(breaks = seq(20, 100, by = 10),
                      limits = c(min(x) - 3, max(x) + 3)) +
  scale_y_continuous(breaks = seq(0, 400, by = 100),
                      limits = c(min(0, min(y)) - 10, max(y) * 1.05)) +
  labs(x = "Test Width (mm)", y = "Total Body Weight (g)",
       title = "Fit Curve: Exponential 3P") +
  theme_bw(base_size = 13, base_family = "sans") +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0),
    panel.grid.major = element_line(color = "grey88", linewidth = 0.3),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "grey40", linewidth = 0.5),
    axis.title = element_text(size = 12, face = "plain"),
    axis.text = element_text(size = 10.5, color = "black"),
    axis.ticks = element_line(color = "grey40")
  )

ggsave("urchin_biomass_plot.png", urchin_biomass, width = 6.2, height = 5.3, dpi = 200)

urchin_biomass
```

## 
